---
title: "Final Writeup"
author: "Sarah Weinstein"
date: "2024-12-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(igraph)
library(networkD3)
library(stringr)
library(terra)
library(sf)
```

1. Introduction:

For our project we intended to create a tool that would help farmers integrate native plants into their crops in order to increase the pollinator density of the area. As we saw in out research that increasing the native plant population in ones crops will lead to an increase in pollinator density(Rao, et al ) We also saw the power of bees in how they can help to spread the growth of plants through pollination as seen with the invasive star thistle(Mciver et al. 2009) Looking at this we can see that having a strong pollinator population is very important to maintain healthy crops and to increase the biodiversity. The interactions between plants and pollinators can help us to learn a lot about the environment that they take place in as we can what plants are favored and which ones are overlooked, as well as seeing what plants bring which bees out. 

Questions:

- Which plant per ecoregion is the most visited?
  * I did not really have any hypothesis for this as I was not sure what plants were preset in each ecoregion but I did hypothesize that the most visited plant would be a native plant. 
  
- What importance does the integration of native plants play in the developing of a healthy and bio diverse cropland?
  * There was really no hypothesis here rather a straight forward answer. It is extremely important as having a biodiverse cropland rich with native plants will encourage pollinators to visit such plants and increase the health of the region. 


Hypothesis: 
for our project we did not really have a hypothesis to test so I am unsure as to what we can do for that aspect of the project. 

2.Data

1) OBA data Spatial data
2) OBA data Network
3) Oregon Ecoregion Data


Tidying the data:
```{r}
oba = read.csv("data/OBA_2018-2023_combined_dataset_working.csv")
## Remove the special character
oba$Associated.plant <- str_replace_all(oba$Associated.plant, "\x92", "")
 
## Remove the rows with no plant name.
oba <- oba[oba$Associated.plant != "",]
 
## Remove "net"
oba <- oba[oba$Associated.plant != "Net",]
 
## Fix yarrow
oba$Associated.plant[oba$Associated.plant == "Yarrow"] <- "Achillea millefolium"
 
## Remove a random weird one
oba <- oba[oba$Associated.plant != "Weedy yellow comp.",]
 
## Remove names in ()
oba$Associated.plant <- str_replace_all(oba$Associated.plant, "\\(.*?\\)", "")
 
## Still some issues with words after commas
oba$Associated.plant <- str_replace_all(oba$Associated.plant, ",.*$", "")
 
## Some have write space at the end of the string now
oba$Associated.plant <- str_replace_all(oba$Associated.plant, "\\s+$", "")
 
## And now there are a few instances where sp doesn't have a period.
oba$Associated.plant <- str_replace_all(oba$Associated.plant, " sp$", " sp.")
 
## Remove the or and everything after it, could also consider dropping these...
oba$Associated.plant <- str_replace_all(oba$Associated.plant, " or .*", "")
 
sort(unique(oba$Associated.plant))[1:10]
```
```{r}
# Count the number of words
oba$plantTaxaWordCount <- str_count(oba$Associated.plant, "\\S+")

# Print the first 50 elements
print(head(oba, 50))
```


```{r}
oba$GenusSpecies <- paste(oba$Genus, oba$Species)
unique_species <- sort(unique(oba$GenusSpecies))
print(unique_species)

num_species <- length(unique_species)
print(paste(num_species))

# Initialize the new column with NA
oba$plantResolution <- NA
oba$beeResolution <- NA
# Loop through each row
for (i in 1:nrow(oba)) {
  # Check if it ends with "aceae" (family)
  if (grepl("aceae$", oba$Associated.plant[i])) {
    oba$plantResolution[i] <- "Family"
  }
  # Check if it has one word and is not a family (genus)
  else if (str_count(oba$Associated.plant[i], "\\S+") == 1) {
    oba$plantResolution[i] <- "Genus"
  }
  # Otherwise, assume it's a species or subspecies
  else {
    oba$plantResolution[i] <- "Species"
  }
}

# Loop through each row
for (i in 1:nrow(oba)) {
  # Check if it has one word and is not a family (genus)
  if (str_count(oba$GenusSpecies[i], "\\S+") == 1) {
    oba$beeResolution[i] <- "Genus"
  }
  # Otherwise, assume it's a species or subspecies
  else {
    oba$beeResolution[i] <- "Species"
  }
}

# Display the distribution of resolutions
table(oba$beeResolution)
# Display the distribution of resolutions
table(oba$plantResolution)
```

ecoregion data
```{r}
OBA_spacial <- oba %>%
  filter(!is.na(Dec..Lat.) & !is.na(Dec..Long.) & !is.na(beeGenus)) %>%
  select(Latitude = Dec..Lat., Longitude = Dec..Long., Plant = plantGenus, Bee = beeGenus)

colnames(OBA_spacial) <- c("Latitude", "Longitude", "Plant", "Bee")

species_sf <- st_as_sf(OBA_spacial, coords = c("Longitude", "Latitude"), crs = 4326)
print(species_sf)
class(species_sf)
```
```{r}
ecoregions <- st_read("data/OR-ecoregions/Ecoregions_OregonConservationStrategy.shp")
```

```{r}
# adding ecoregion column to oba data
template <- rast(vect(ecoregions))

rast_ecoregions <- rasterize(vect(ecoregions), template, field = "Ecoregion")
rast_ecoregions <- project(rast_ecoregions, crs(species_sf))
 #extracts ecoregion data for every point
species_sf$Ecoregion <- terra::extract(rast_ecoregions, species_sf, ID = FALSE)[[1]]

species_sf <- species_sf %>% filter(!is.na(Ecoregion))

unique(species_sf$Ecoregion)
```


Visualization:

Sankey plots 
```{r}
library(dplyr)
library(igraph)

# Example species_sf structure
# species_sf contains columns: Ecoregion, Plant, Bee
# Example structure: species_sf <- data.frame(Ecoregion, Plant, Bee)

# Group by Ecoregion and create edge lists
edge_lists <- species_sf %>%
  filter(!is.na(Plant), !is.na(Bee)) %>% # Ensure no NA values
  group_by(Ecoregion) %>%
  summarise(edges = list(data.frame(Plant, Bee)), .groups = "drop")

# Create a network for each Ecoregion
networks <- lapply(edge_lists$edges, function(edges) {
  graph_from_data_frame(edges, directed = FALSE)
})

# Name networks by Ecoregion
names(networks) <- edge_lists$Ecoregion

# Example: Analyze or plot the network for a specific Ecoregion
plot(networks[[1]], main = paste("Network for Ecoregion:", names(networks)[1]))

```
```{r}
library(dplyr)
library(networkD3)

# Group by Ecoregion and prepare Sankey input
sankey_data <- species_sf %>%
  filter(!is.na(Plant), !is.na(Bee)) %>% # Remove rows with NA values
  group_by(Ecoregion) %>%
  summarise(
    nodes = list(unique(c(Plant, Bee))), # Combine unique nodes (Plants and Bees)
    links = list(data.frame(source = Plant, target = Bee)), # Create links
    .groups = "drop"
  )

# Create and display Sankey diagrams for each Ecoregion
sankey_plots <- lapply(1:nrow(sankey_data), function(i) {
  # Extract nodes and links
  ecoregion <- sankey_data$Ecoregion[i]
  node_list <- unique(c(sankey_data$nodes[[i]]))
  link_df <- sankey_data$links[[i]]

  # Create node data frame with indices
  nodes <- data.frame(name = node_list)

  # Convert links to indices and add a Value column
  links <- link_df %>%
    mutate(
      source = match(source, nodes$name) - 1, # Convert to zero-based index
      target = match(target, nodes$name) - 1,
      value = 1 # Assign a default value of 1 for all links
    )

  # Create Sankey diagram
  sankeyNetwork(
    Links = links,
    Nodes = nodes,
    Source = "source",
    Target = "target",
    Value = "value", # Use the Value column created above
    NodeID = "name",
    units = "Interactions",
    fontSize = 12,
    nodeWidth = 30
  )
})

# Display Sankey diagram for the first Ecoregion as an example
sankey_plots[[1]] # Replace [1] with any index to display another Ecoregion
sankey_plots[[2]]
sankey_plots[[3]]
sankey_plots[[4]]
sankey_plots[[5]]
sankey_plots[[6]]
sankey_plots[[7]]
sankey_plots[[8]]
```

Plant Pollinator plots 
```{r}
library(dplyr)
library(ggplot2)

# Example species_sf structure
# species_sf contains columns: Ecoregion, Plant, Bee

# Summarize the 5 most common bees for each ecoregion
top_bees <- species_sf %>%
  filter(!is.na(Bee)) %>% # Ensure no NA values
  group_by(Ecoregion, Bee) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Ecoregion) %>%
  slice_max(Count, n = 5, with_ties = FALSE) %>%
  ungroup()

# Summarize the 5 most common plants for each ecoregion
top_plants <- species_sf %>%
  filter(!is.na(Plant)) %>% # Ensure no NA values
  group_by(Ecoregion, Plant) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Ecoregion) %>%
  slice_max(Count, n = 5, with_ties = FALSE) %>%
  ungroup()

# Plot for bees
bee_plots <- top_bees %>%
  group_split(Ecoregion) %>%
  lapply(function(data) {
    ggplot(data, aes(x = reorder(Bee, -Count), y = Count, fill = Bee)) +
      geom_bar(stat = "identity") +
      coord_flip() +
      labs(
        title = paste("Top 5 Bees in Ecoregion:", unique(data$Ecoregion)),
        x = "Bee",
        y = "Count"
      ) +
      theme_minimal()
  })

# Plot for plants
plant_plots <- top_plants %>%
  group_split(Ecoregion) %>%
  lapply(function(data) {
    ggplot(data, aes(x = reorder(Plant, -Count), y = Count, fill = Plant)) +
      geom_bar(stat = "identity") +
      coord_flip() +
      labs(
        title = paste("Top 5 Plants in Ecoregion:", unique(data$Ecoregion)),
        x = "Plant",
        y = "Count"
      ) +
      theme_minimal()
  })

# Example: Display the first barplot for bees and plants
bee_plots[[1]] # Replace [1] with other indices for different ecoregions
plant_plots[[1]] # Replace [1] with other indices for different ecoregions
bee_plots[[2]] # Replace [1] with other indices for different ecoregions
plant_plots[[2]]
bee_plots[[3]] # Replace [1] with other indices for different ecoregions
plant_plots[[3]]
bee_plots[[4]] # Replace [1] with other indices for different ecoregions
plant_plots[[4]]
bee_plots[[5]] # Replace [1] with other indices for different ecoregions
plant_plots[[5]]
bee_plots[[6]] # Replace [1] with other indices for different ecoregions
plant_plots[[6]]
bee_plots[[7]] # Replace [1] with other indices for different ecoregions
plant_plots[[7]]
bee_plots[[8]] # Replace [1] with other indices for different ecoregions
plant_plots[[8]]
```


Map
```{r}
or <- map_data("county", "oregon") %>% 
  select(lon = long, lat, group, id = subregion)
library(ggspatial)
ggplot() +
  geom_polygon(data = or, aes(x = lon, y = lat, group = group), fill = "lightgrey", color = "black") +
  geom_sf(data = species_sf, aes(color = Ecoregion), size = 3) +
  theme_minimal() +
  labs(
    title = "Bee Species by Ecoregions",
    x = "Longitude",
    y = "Latitude",
  ) +
  annotation_scale(location = "br", width_hint = 0.2) +
  annotation_north_arrow(location = "tr", width = unit(.5, "in")) +
  theme(legend.position = "none")

```

















Visualization:
```{r}

#what we did here was we created graphs that show the top bees and plants visited per ecooregin and made bar graphs of them 
library(dplyr)
library(ggplot2)
# Example species_sf structure
# species_sf contains columns: Ecoregion, Plant, Bee

# Summarize the 5 most common bees for each ecoregion
top_bees <- species_sf %>%
  filter(!is.na(Bee)) %>% # Ensure no NA values
  group_by(Ecoregion, Bee) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Ecoregion) %>%
  slice_max(Count, n = 5, with_ties = FALSE) %>%
  ungroup()

# Summarize the 5 most common plants for each ecoregion
top_plants <- species_sf %>%
  filter(!is.na(Plant)) %>% # Ensure no NA values
  group_by(Ecoregion, Plant) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Ecoregion) %>%
  slice_max(Count, n = 5, with_ties = FALSE) %>%
  ungroup()

# Plot for bees
bee_plots <- top_bees %>%
  group_split(Ecoregion) %>%
  lapply(function(data) {
    ggplot(data, aes(x = reorder(Bee, -Count), y = Count, fill = Bee)) +
      geom_bar(stat = "identity") +
      coord_flip() +
      labs(
        title = paste("Top 5 Bees in Ecoregion:", unique(data$Ecoregion)),
        x = "Bee",
        y = "Count"
      ) +
      theme_minimal()
  })

# Plot for plants
plant_plots <- top_plants %>%
  group_split(Ecoregion) %>%
  lapply(function(data) {
    ggplot(data, aes(x = reorder(Plant, -Count), y = Count, fill = Plant)) +
      geom_bar(stat = "identity") +
      coord_flip() +
      labs(
        title = paste("Top 5 Plants in Ecoregion:", unique(data$Ecoregion)),
        x = "Plant",
        y = "Count"
      ) +
      theme_minimal()
  })

# Example: Display the first barplot for bees and plants
bee_plots[[1]] # Replace [1] with other indices for different ecoregions
plant_plots[[1]] # Replace [1] with other indices for different ecoregions
bee_plots[[2]] # Replace [1] with other indices for different ecoregions
plant_plots[[2]]
bee_plots[[3]] # Replace [1] with other indices for different ecoregions
plant_plots[[3]]
bee_plots[[4]] # Replace [1] with other indices for different ecoregions
plant_plots[[4]]
bee_plots[[5]] # Replace [1] with other indices for different ecoregions
plant_plots[[5]]
bee_plots[[6]] # Replace [1] with other indices for different ecoregions
plant_plots[[6]]
bee_plots[[7]] # Replace [1] with other indices for different ecoregions
plant_plots[[7]]
bee_plots[[8]] # Replace [1] with other indices for different ecoregions
plant_plots[[8]]
```